Responsive GPU Workgroup Management in the Presence of Persistent Kernels

### 2016-11-08 write-ups/PLDI/paper.tex
Started on paper intro based on notes with Tyler.

>>>>

\ADComment{Be sure to say that a GPU \emph{kernel} is.}

Although GPU acceleration of general-purpose computation has focused
on regular data-parallel algorithms, many data-parallel algorithms are
\emph{irregular} \ADComment{say what that means}.  There is growing
interest in accelerating irregular data parallel algorithms on GPUs,
including \ADComment{name some algorithms and say why they matter},
with several recent notable successes~\cite{...}.

By their very nature, irregular algorithms demand synchronization
between the workgroups that execute a GPU kernel.  For example, work
stealing algorithms require the ability for one workgroup to steal
from the queue of another workgroup, which can be safely achieved by
protecting the queue of each workgroup using a mutex.  Breadth-first
graph traversal uses a frontier-based approach that requires a barrier
synchronization between all workgroups to allow computation to proceed
from one frontier to the next.  Inter-workgroup synchronization
involves \emph{blocking} between workgroups, e.g.\ workgroup A may be
blocked until workgroup B releases a mutex, or blocked until all other
workgroups arrive at an inter-workgroup barrier.  Thus, irregular data
parallel algorithms that rely on inter-workgroup synchronization are
\emph{blocking algorithms}.\footnote{\ADComment{Perhaps make the point
    that inter-workgroup synchronization is possible in very rare
    cases without blocking, e.g. threadfence reduction.}}

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular data
parallel algorithms accelerated on a GPU, this translates to requiring
fair scheduling of workgroups.  Without fairness, a workgroup may fail
to make progress, leading to starvation.  For example, if workgroup B
holds a mutex and workgroup A is spinning, waiting for the mutex to be
released, an unfair scheduler can forever schedule A, so that B never
releases a mutex.  Similarly, an unfair scheduler can induce
starvation if workgroup A is at an inter-workgroup barrier, spinning
until other workgroups arrive.

Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
and HSA~\cite{...}---specify very few constraints on the manner in
which workgroups are scheduled.  In particular, they do \emph{not}
guarantee fair scheduling, and fair scheduling is not observed in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.  If a kernel is executed
with more workgroups than there are compute units then at least two
workgroups must share a single compute unit.  The simplest way to
handle this is an \emph{occupancy-bound} execution model, whereby once
a workgroup has commenced execution on a compute unit (is has become
\emph{occupant}), the workgroup has exclusive access to the compute
unit until it has finished execution, after which a subsequent
workgroup can be bound to the compute unit.  There is empirical
evidence that the occupancy-bound execution model is widely
implemented in practice~\cite{...}.

Clearly, the occupancy-bound execution model does not provide fair
scheduling between workgroups in general: if all compute units are
occupied then a not-yet-occupant workgroup will not be scheduled until
some occupant workgroup has completed execution.  Yet the execution
model \emph{does} provide fair scheduling between occupant workgroups,
which are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume this occupancy-bound execution model, exploiting it by
scheduling no more workgroups than there are available compute units.
Because current GPUs do provide the occupancy-bound execution model in
practice, this strategy allows starvation-free blocking
synchronization between workgroups.

However, none of the current GPU programming models mandate the
occupancy-bound execution model, and there are clear reasons why this
model is not actually desirable from the point-of-view of GPU vendors
nor end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all workgroups, the GPU cannot be used for other
tasks, such as graphics rendering or shorter, higher-priority
computations, in the meantime.  Precisely because of this problem,
operating systems employ a GPU \emph{watchdog} to heuristically detect
and kill long-running computations on the GPU that is being used for
rendering.  If one disables this watchdog and launches a long-running
kernel, the display becomes unresponsive for the duration of kernel
execution.  Second, \emph{energy throttling} is an important concern
for energy-constrained devices, such as smartphones and tablets.  In
the context of mobile GPU computing, it is desirable for a GPU driver
to be able to power down one or more compute units if the battery
level is low or if the user puts the device into a low power state.
Implementing energy throttling requires being able to suspend the
execution of a workgroup, power down the compute unit to which the
workgroup was bound, and schedule the workgroup for later execution on
one of the other compute units when it becomes free.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that GPU
vendors are reluctant to commit to the occupancy-bound execution model
that they currently implement, due to the need for multi-tasking and
energy throttling, yet do not wish to guarantee fair scheduling, due
to the runtime overhead associated with supporting full preemption of
workgroups, e.g.\ based on time-slicing.  Vendors instead wish to
retain the essence of the occupancy-bound execution model, which is
simple to implement efficiently, providing preemption for the special
cases associated with multi-tasking for high-priority jobs (such as
graphics) and energy throttling.  In particular, they do not want to
commit to fair scheduling of a workgroup once it has been preempted.
As argued above, without fairness it is not possible to implement
non-blocking algorithms, including data parallel irregular algorithms,
safely.

  Doesn't fascilitate energy throttling or
  multi-tasking. Multi-tasking is important esp. for responsive
  systems. (Say something colourful here, its an enabler. A use case
  for today and many use cases for tomorrow.)

  Our Proposal: Non-blocking GPU programs (majority of kernels today)
  can be exucuted however the GPU scheduler sees fit. Algorithm that
  require a degree of fairness (e.g. blocking algorithms) must be
  written a small set of language extensions, and launched in a
  cooperative mode. This indicates that the scheduler must interact
  with these kernels in a special manner via the language
  extensionts. These kernels are written with language extensions that
  allow designated points at which a workgroup can be de-scheduled or
  re-scheduled to be identified.

  In traditonal co-operative multi-tasking, a task reliquishes control
  and regains control in a stateful manner. We propose a lighter
  weight model we believe to suitible for the type of blocking
  algorithms we are aware of for GPUs. At a point where the algorithm
  could proceed with fewer workgroups, a workgroup can offer itself
  to be killed. This indicates to the scheduler that the algorithm
  would happily proceed with the workgroup continuing, but is prepared
  for the workgroup to be destroyed for good. In addition, the programmer
  identifies designated points where it would be beneificial to join the
  computation if available. On offering to fork, the algorithm indicates
  that it is prepared to proceed with the existing workgroups, but is able
  to take advantage of one or more additional workgroups commencing at the
  given program point. Inheriting the state of the invoking workgroup.

  The scheduler agrees to scheduler the workgroups fairly, but is at
  liberty to eliminate workgroups for good when they offer themselves
  to be killed. This allows the schduler to repond to changing depands
  on the GPU, e.g. to implement energy throttling.

  *Note: its not really completely fair, its occupancy bound

  This programming model creates a contract between the scheduler
  and cooperative kernel:

  Functional requirements: Scheduler agrees that cooperative kernels
  have fairly scheduled workgroups. The Cooperative kernel is written
  in a way that can tolerate threads leaving or joining at the
  designated areas (e.g. a thread should not opt to be killed if it
  contains local work to do, and a programmer should be prepared for
  the number of executing workgroups to change dynamically during
  computation).

  Non-Functional: The cooperative kernel must ensure that offer_kill
  is executed frequently enough such that the system can accomodate
  soft-real time constraints (e.g. allowing a smooth framerate for
  graphics or throttling the energy before the mobile battery drains).
  To enable the cooperative execute efficiently, the scheduler should
  adopt a policy that avoids under utililasion of hardware resources
  for the cooperative kernel (e.g. the scheduler should not kill
  workgroups unless they are required).

  The programming model is backwards compatible because semantically
  it is fine for these things to be ignored and GPUs already give
  occupant bound fairness.

  Our contributions:

  - Programming model in detail (semantics of new operations)

  - Prototype implementation of the programming model using OpenCL 2.0
    advance features to build the scheudler and implement a clang
    based compiler to transform input programs using the language
    extensions to lower-level programs that interact with the
    scheduler. Our prototype simulates multiple independent kernels
    into a mega-kernel. A vendor specific implementation could achieve
    higher efficiency through modications to the GPU driver that may
    take advantage of hardware features. (e.g. mega kernel might not
    be good to compile). Its reasonable to think our proof of concept
    provides a lower bound on how well we can do.

  - We ported a large set of blocking algorithms to our programming
    model. We evaluate co-operative scheduling with a cooperative
    kernel and non-coperative kernel. GPUs from multiple vendors and
    access various things.

  - Our results are promising.

  - In the evaluation, make a note that the applications we worked on
    Namely if they use only a barrier or work-stealing idiom, are
    straightforward to adopt to our programming model.
